---
title: "Generating A Word Cloud"
description: |
  This post is about generating a word cloud. Though you can easily generate one using numerous online programs, utilizing R allows for greater flexibility as well as improving your knowledge of the R Language
author:
  - name: Jonathan Choo
    Affiliation: University of Nottingham, Malaysia
date: 2023-02-22
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(repos = list(CRAN="http://cran.rstudio.com/"))
```

To start, we first need to install the following packages and load it to our R session

```{r Load Packages, echo=TRUE, results='hide'}
install.packages("wordcloud")
library(wordcloud)
install.packages("RColorBrewer")
library(RColorBrewer)
install.packages("wordcloud2")
library(wordcloud2)
install.packages("dplyr")
library(dplyr)
```

Next, we need to store our text data as a vector so that we can manipulate it easier. To do this, we use the *corpus()* function from the *tm* package to create a corpus of our text data

```{r echo=TRUE, results='hide'}
install.packages("tm")
library(tm)
```

Then, load up your data into your working directory and workspace. Lets call the dataframe as **dataframe**

```{r ActualRead, include=FALSE}

dataframe <- read.csv("..\\2023-02-22-generating-a-word-cloud\\data.csv")
dataframe <- dataframe[-c(1,2),]
```

```{r TextRead, eval=FALSE, echo=TRUE}

dataframe <- read.csv("your_file.csv")

```

What you want to do next is to store the text data from your dataframe into a vector and create a corpus *(let's call this corpus as docs)* out of that vector. In this example, Q19 contains all the text data information that we want to turn into a word cloud

```{r echo=TRUE}
#Create a vector containing only the text. 
textdata <- dataframe$Q19

# Create a corpus  
docs <- Corpus(VectorSource(textdata))
```

Next, we want to make sure that only important/useful words are included in the corpus. Currently, numbers, punctuations and stopwords (e.g., and, the, a, is) are still included in the corpus. We need to clean these up or they will show up in the word cloud. To do so, we use the *tm_map* function from the *tm* package

```{r echo=TRUE}
docs <- docs %>%
  tm_map(removeNumbers) %>%
  tm_map(removePunctuation) %>%
  tm_map(stripWhitespace)
docs <- tm_map(docs, content_transformer(tolower))
docs <- tm_map(docs, removeWords, stopwords("english"))
```

As we are working with bilingual answers (English and Bahasa Malaysia), we also want to remove the BM stopwords. This requires data from the *malaytextr* package

```{r echo=TRUE, results='hide'}
install.packages("malaytextr")
library(malaytextr)

#We store the stopwords as a matrix in an object called stopmalay
stopmalay <- as.matrix(malaytextr::malaystopwords)

#We then remove the malay stopwords from the earlier document
docs <- tm_map(docs, removeWords, stopmalay)
```

What you want to do as a next step is to have a new dataframe containing each word in your first column and their frequency in the second column.

This can be done by creating a document term matrix with the *TermDocumentMatrix* function from the *tm* package.

```{r echo=TRUE}
dtm <- TermDocumentMatrix(docs) 
matrix <- as.matrix(dtm) 
words <- sort(rowSums(matrix),decreasing=TRUE) 
df <- data.frame(word = names(words),freq=words)
```

Finally, we generate our word cloud using the *wordcloud* package

```{r echo=TRUE}
set.seed(1234) # for reproducibility 

wordcloud(words = df$word, freq = df$freq, 
          min.freq = 1, 
          max.words=200, 
          random.order=FALSE,
          rot.per=0.35,
          scale=c(2.5,0.25),
          colors=brewer.pal(8, "Dark2"))
```
Sometimes, depending on the size of your corpus, the plot may not show all words. If this is the case, you should play around with the *scale=c(2.5,0.25)* argument until an acceptable number of information is present in your word cloud.

Eagle-eyed individuals will see that we have not used the *wordcloud2* package. This package is a bit more fun to use, allowing us to create word clouds in various shapes such as below:

```{r echo=TRUE}
wordcloud2(data=df, size = 0.4, shape = 'diamond')
```

Happy Learning!

Jonathan
